---
title: "Problem Set N°1"
author: "Valentina Andrade"
abstract: "The following report contains the exercises requested in problem set 1. In the first part you can download the proofs of some properties and/or results related to AR, MA and ARMA process. In the second part, the Box-Jenkins methodology is applied to study three series of the Chilean economy: inflation, exchange rate and IPSA. One of the most important results of both exercises is related to how to apprehend time series structures, either theoretically or empirically we can say something that Wold 's theorem had already anticipated '*'Any stationary series can beexpressed as the sum of two components: a perfectly forecastable series and a moving average of possily infinite order'*'"

date: August 27, 2011
output:
  prettydoc::html_pretty:
    theme: tactile
    highlight: github
    math: katex

---

# Part 1 

[Download document (proofs 1 to 6)](https://drive.google.com/drive/u/0/folders/1RZOutpun_CCGriqZta94c639J7ycveYe)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, dev = 'svg', message = FALSE)
library(tidyverse)
library(lubridate)
library(patchwork)
library(forecast) # autoplot, auto.arima
library(tseries) # Dickey-Fuller
library(ggfortify)
library(xts)
library(readxl)
library(showtext)

theme_set(theme_minimal(base_size = 12))
theme_set(theme_minimal(base_size = 14))
font_add_google("IBM Plex Sans", "ibm")
showtext_auto()
```

```{r, funciones, include=FALSE}
get_time_series <- function(data, time, value){
  # https://stackoverflow.com/questions/11422493/using-r-how-do-i-create-a-time-series-object-with-dates 
  xts(
    x = pull(data, value),
    order.by = pull(data, time),
    frequency = 12
  ) 
}

tidy_adf_test <- function(serie){
  # ?adf.test
  # https://www.r-bloggers.com/2021/12/augmented-dickey-fuller-adf-test-in-r/
  # https://rstudio-pubs-static.s3.amazonaws.com/843574_4aec57ae7a0d43ff8d0c152729f0ed2c.html#prueba-de-ra%C3%ADz-unitaria-de-dickey-fuller
  
  serie  %>%  
    na.omit() %>% 
    adf.test() %>% 
    broom::tidy() %>% 
    mutate(
      `resultado 95%` = ifelse(
        p.value < 0.05,
        "Es I(0),  no unit-root",
        "Existe unit-root")
    ) %>% 
    rename(`Valor-p` = p.value) %>% 
    select(method, `Valor-p`, everything())
}

autoplot_acf_pact <- function(serie){
  
  p1 <- ggAcf(serie) + labs(title = NULL)
  p2 <- ggPacf(serie) + labs(title = NULL)
  
  p1 + p2
  
}

modelos_arma <- function(serie, ar_orders = 1:6, ma_orders = 1:6){
  
  # test nombre_colmm <- "infsv_sa"
  
  modelos <- crossing(ar = ar_orders, ma = ma_orders) %>% 
    pmap_df(function(ar = 6, ma = 3){
      
      cli::cli_alert_info("Ajustando ARMA({ar}, {ma})")
      
      ajuste <- Arima(serie, order = c(ar, 0, ma), include.mean = TRUE, method = "ML")
      
      # autoplot(serie)
      # autoplot(ajuste$residuals)
      
      # ggAcf(serie)
      # ggAcf(ajuste$residuals)
      
      # tsdiag(ajuste)
      
      output <- broom::glance(ajuste)
      
      output <- select(output, -nobs)
      
      output
      
      Box.test(serie, type = "Ljung-Box")
      Box.test(ajuste$residuals, type = "Ljung-Box")
      
      # tsdiag(ajuste)
      
      test_residuos <- Box.test(ajuste$residuals, type = "Ljung-Box")
      test_residuos <- broom::glance(test_residuos)
      
      output <- output %>% 
        add_column(`Box-Ljung test residuos p value` = test_residuos$p.value) 
      
      output <- output %>% 
        mutate(`Ajuste` = str_glue("ARMA({ar}, {ma})"), .before = 1)
      
      output
      
    })
  
  modelos %>% arrange(desc(`Box-Ljung test residuos p value`))
  modelos %>% arrange(sigma)
  modelos %>% arrange(desc(logLik))
  modelos %>% arrange(desc(BIC))
  modelos %>% arrange(desc(AIC))
  
  modelos <- modelos %>%
    arrange(desc(`Box-Ljung test residuos p value`))
  
  auto.arima(serie, max.p = 6, max.q = 6, stepwise = FALSE)
  
  modelos
  
}

modelo_summary <- function(fit){
  
  bind_cols(broom::tidy(fit), confint(fit))
  
}

plot_ajuste <- function(modelo){
  
  modelo$x %>% 
    as.data.frame() %>% 
    set_names("serie") %>% 
    rownames_to_column("periodo") %>% 
    as_tibble() %>% 
    mutate(
      periodo = ymd(periodo),
      ajuste = fitted(modelo)
      ) %>% 
    pivot_longer(cols = -periodo) %>% 
    
    ggplot() +
    geom_line(aes(periodo, value, color = name)) +
    scale_color_viridis_d(begin = 0.3, end = 0.8)
    
}
```

# Part 2

ARMA models have been presented as a parsimonious tool to describe non-stationary stochastic processes. In theory, non-stationary series can be represented by an MA($\infty$), i.e., capturing the entire memory of the series. 

In practice this is very expensive, so we will show how we can approximate an MA($\infty$) from an ARMA($p,q$) model, with few parameters (i.e. $p+q$ is small).  We will be guided by the methodology of *Box and Jenkins* to achieve this task. 

1. In order to use ARMA we need the non-stationary components or "trends around the mean" or "trends around the variance" to be removed. In addition to using transformations, we test a unit root test (Dickey Fuller's test).

2. Other deterministic components are removed. In our case this is important because before 2001 we find that there is a clear inflationary path, and that this is evidently due to the change in the monetary policy regime (3% rule). 

3. Third, we compute ACF and PACF to identify the order and type of the underlying model.  

4. The model is estimated assuming the proposed model with p and q. 

5. Identification tests are performed and the adequacy of the identification is evaluated. In this report we give importance to AIC and Ljung-Box.

6. In-sample predictions of the estimated model are made.

## Data exploration

```{r data}
datos <- readxl::read_excel(here::here("input/datos.xlsx"))
datos <- datos %>%  select(-inf,-inf_sa,-infsv, -ipsa, -tcn)
datos <- janitor::clean_names(datos)

datos <- datos %>%
  mutate(periodo = as.Date(periodo))

datos_long <- pivot_longer(datos, cols = -periodo)

ggplot(datos_long) +
  geom_vline(
    aes(xintercept = fecha),
    data = tibble(fecha = ymd("2000-01-01", "2020-01-01")),
    color = "darkred",
    size = 1.2
    ) + 
  geom_line(aes(periodo, value), color = "gray60") + labs(y = "") + 
  facet_wrap(vars(name), scales = "free_y", nrow = 2) 

```


*Figure 1*. Series `infsv_sa` (IPC), `IPSA_sa` (IPSA), `tcn_sa` (Exchange Rate CLP/USD) (1990- 2022)

Pese a que las series utilizadas son las desestacionalizadas, as can be shown in *Figure 1*, inflation (measured by the consumer price index) presents a clear trend before 2001. This price growth trend was stabilized after the Central Bank set a target of around 3% inflation and a policy of nominalization [(Fuentes et al, 2003)](https://si2.bcentral.cl/public/pdf/documentos-trabajo/pdf/dtbc197.pdf). Similarly, since 2020, due to the health crisis caused by the COVID-19 pandemic, the consequences have also been reflected in an increase in the cost of living. 

In order to isolate the trends mentioned above, we have chosen to limit the period of analysis from 2001 to 2020, both for inflation and for the other variables of interest, in order to make the models more comparable. We will use the series shown in *Figure 2* for the following steps.   


```{r}
datos_2020 <- datos %>% 
  filter(between(periodo, ymd("2000-01-01"), ymd("2020-01-01")))

datos_2020_long <- datos_long %>% 
  filter(between(periodo, ymd("2000-01-01"), ymd("2020-01-01")))


ggplot(datos_2020_long) +
  geom_line(aes(periodo, value), color = "gray60") +
  facet_wrap(vars(name), scales = "free_y", nrow = 2)

```

*Figure 2*. Series `infsv_sa` *(Inflation)*, `ipsa_sa` *(IPSA)*, `tcn_sa` *(Exchange Rate CLP/USD)* (2001- 2022)


## 1. `Inflation`

```{r}
nombre_colmm <- "infsv_sa"
nombre_serie <- "Inflation %"

serie <- get_time_series(datos_2020, time = "periodo", value = nombre_colmm)

autoplot(serie) + 
  labs(x = "Date", y = nombre_serie, caption = "Figure 3")

```

First, in Figure 3 we have a clear representation of an increasing trend in the price level. As we mentioned at the beginning, ARMA models work on the basis of non-stationary series, but graphically it seems that inflation still has a trend component. We will use the Dickey Fuller unit root test to conjecture if there is evidence of this trend. Formally, 

$$\triangle Y_t = \alpha + \phi y_{t-1}+ \varepsilon$$

$H_0: \phi = 0\Rightarrow$ Presence of stochastic trend in the observations.

$H_1: \phi <0 \Rightarrow$: No presence of stochastic trend in the observations.

```{r}
knitr::kable(tidy_adf_test(serie), caption = "Table 1. Dickey-Fuller Test for Inflation series")
```

Table 1 shows that with 95% confidence, we cannot reject the null hypothesis. That is, it is likely to say that there is a stochastic trend in this series. Our calculations show that it is a trend in means so it can be solved with a simple differencing (if it were a trend in variances a logarithmic transformation would be appropriate). After the transformation we plot the series in *Figure 4*.Table 2 shows that we can now reject the null hypothesis with 95% confidence.
```{r}
serie <- diff(serie)

autoplot(serie) + 
  labs(x = "Fecha", y = nombre_serie, caption = "Figure 4.")

knitr::kable(tidy_adf_test(serie), caption = "Table 2. Dickey-Fuller Test for diff(inflation) ")

```

We will now explore the order of the AR and MA processes.  On the one hand, the ACF gives us information about the order $q$ of the MA. The figure is not very clear about whether the value is at 1 or much higher (there are values near to 14). On the other hand, the (partial) PACF gives us the p-value, i.e., the order of the AR(p) process. The figure shows with much more certainty that the process "dies" between 3 and 5. Evidently the value 5 could be possible only because of a convenience of the size of the interval.


```{r}
autoplot_acf_pact(serie) + 
  plot_annotation(title = str_glue("Figure 5. {nombre_serie}: ACF y PACF"))

modelos_arma(serie, ar_orders = 1:5, ma_orders = 1:5) %>% 
  arrange(AIC) %>% 
  head(12) %>% 
  knitr::kable(., caption = "Table 3. ARMA(p,q) Iteration")

AR <- 3
MA <- 4

modelo <- Arima(serie,  order = c(AR, 0, MA), include.mean = TRUE)
```

A function has been created to order the models according to their fit considering AIC (information criterion), Box-Ljung which studies that any series of autocorrelations is non-zero (Portmanteau test), logLik. Taking this information, the function penalizes the ARMAs that have higher order p+q. That is why  we select the model `r  str_glue("Modelo ARMA({AR}, {MA})")`which
has AIC of `r modelo$aic`. The estimated parameters are:

```{r}
knitr::kable(modelo_summary(modelo) %>% filter(estimate != 0))
```

Auto-correlation functions of residuals are represented in ACF. As can be seen, the correlogram "dies" at zero so it evidently reveals to be white noise. This tells us that the residuals have no structure and therefore the model has been well specified and does not store information about the series.  

```{r}
ggtsdiag(modelo)
```

The Ljung Box statistical significance gives us a robustness test: autocorrelation does not occur for any lag of the series (see order equal to 10).


```{r}
plot_ajuste(modelo)
```


The last step of Box-Jenkins corresponds to prediction. As we can see in the figure presented, the values predicted by the ARMA model follow very closely the empirical series.

## `IPSA`

```{r}
nombre_colmm <- "ipsa_sa"
nombre_serie <- "El nombre bonito"

serie <- get_time_series(datos_2020, time = "periodo", value = nombre_colmm)

autoplot(serie) + 
  labs(x = "Fecha", y = nombre_serie)

knitr::kable(tidy_adf_test(serie))

# serie <- diff(serie)
# 
# autoplot(serie) +
#   labs(x = "Fecha", y = nombre_serie)
# 
# knitr::kable(tidy_adf_test(serie))

autoplot_acf_pact(serie) + 
  plot_annotation(title = str_glue("{nombre_serie}: ACF y PACF"))

modelos_arma(serie, ar_orders = 1:6, ma_orders = 1:6) %>% 
  arrange(AIC) %>% 
  head(12) %>% 
  knitr::kable()

AR <- 6
MA <- 6

modelo <- Arima(serie,  order = c(AR, 0, MA), include.mean = TRUE)
```

We select the model `r  str_glue("Modelo ARMA({AR}, {MA})")`, el cual
posee AIC de `r modelo$aic`. The estimated parameters are:


```{r}
knitr::kable(modelo_summary(modelo) %>% filter(estimate != 0))
```

Auto-correlation functions of residuals are represented in ACF and PACF, to figure the order of MA and AR models, respectively.

```{r}
autoplot_acf_pact(modelo$residuals) +
  plot_annotation(title = str_glue("Residuos modelo ARMA({AR}, {MA}) {nombre_serie}: ACF y PACF"))
```

```{r}
ggtsdiag(modelo, gof.lag = 24)
```

```{r}
plot_ajuste(modelo)
```


## `3. Exchange Rate`

```{r}
nombre_colmm <- "tcn_sa"
nombre_serie <- "El nombre bonito"

serie <- get_time_series(datos_2020, time = "periodo", value = nombre_colmm)

autoplot(serie) + 
  labs(x = "Fecha", y = nombre_serie)

knitr::kable(tidy_adf_test(serie))

# serie <- diff(serie)
# 
# autoplot(serie) +
#   labs(x = "Fecha", y = nombre_serie)
# 
# knitr::kable(tidy_adf_test(serie))

autoplot_acf_pact(serie) + 
  plot_annotation(title = str_glue("{nombre_serie}: ACF y PACF"))

modelos_arma(serie, ar_orders = 1:6, ma_orders = 1:6) %>% 
  arrange(AIC) %>% 
  head(12) %>% 
  knitr::kable()

AR <- 3
MA <- 2

modelo <- Arima(serie,  order = c(AR, 0, MA), include.mean = TRUE)

# auto.arima(serie)

```

We select the model `r  str_glue("Modelo ARMA({AR}, {MA})")`, el cual
posee AIC de `r modelo$aic`. The estimated parameters are:


```{r}
knitr::kable(modelo_summary(modelo) %>% filter(estimate != 0))
```

Auto-correlation functions of residuals are represented in ACF and PACF, to figure the order of MA and AR models, respectively.

```{r}
autoplot_acf_pact(modelo$residuals) +
  plot_annotation(title = str_glue("Residuos modelo ARMA({AR}, {MA}) {nombre_serie}: ACF y PACF"))
```

```{r}
ggtsdiag(modelo, gof.lag = 24)
```

```{r}
plot_ajuste(modelo)
```

# References

[*Hamilton, J. D. (1994). Time series analysis. Princeton University Press. Chapters 1-3*](https://books.google.cl/books?hl=es&lr=&id=BeryDwAAQBAJ&oi=fnd&pg=PP1&dq=hamilton+time+series+analysis&ots=BgCP8_bUhj&sig=3zy12_T3kgYPWPja8y5F1lJPZWQ&redir_esc=y#v=onepage&q=hamilton%20time%20series%20analysis&f=false)

[*Brockwell, P. J., & Davis, R. A. (1991). Time Series: Theory and Methods.. Chapter 3*](https://books.google.cl/books?hl=es&lr=&id=TVIpBgAAQBAJ&oi=fnd&pg=PA1&dq=brockwell+time+series+theory+and+methods&ots=nl2yfymmW6&sig=ktnul2bhi-vKD0GpvtG7KJIbol8&redir_esc=y#v=onepage&q=brockwell%20time%20series%20theory%20and%20methods&f=false)

[*Hayashi, F. (2011). Econometrics. Princeton University Press.*](https://books.google.cl/books?hl=es&lr=&id=QyIW8WUIyzcC&oi=fnd&pg=PP2&dq=hayashi+econometrics&ots=SFv20wuaK4&sig=kS7vf0fprmxEkGeJriM1xuX2B44&redir_esc=y#v=onepage&q=hayashi%20econometrics&f=false)


